---
title: "REMIND/IAM Data Analysis Using quitte"
author: "Michaja Pehl"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quitte-data-analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	collapse = TRUE,
	comment = "#>",
	fig.height = 5, 
	fig.width = 7
)
```

The `quitte` package is a grab-bag of utility functions to work with REMIND/IAM 
results (or any kind of data) in a `.mif`-like format. It builds on the 
capabilities of the `dplyr`, `tidyr`, and `ggplot2` packages.

# Load required packages
```{r load necessary packages}
library(tidyverse)
library(quitte)
```

Generally, when working with quitte, you will need the `dplyr` and `tidyr` 
packages. `tidyverse` is a place holder package loading these two, as well as 
the `ggplot2` package and a couple of others.


# Data Wrangling

As `quitte` builds on the capabilities and practices of the _tidy data_ 
packages, it is highly recommended to read [this excellent tutorial introducing 
`dplyr` and `tidyr`
](https://datacarpentry.org/R-ecology-lesson/03-dplyr.html). You 
should be comfortable with the concept of piping (`%>%`), as well as the 
`filter()`, `select()`, `group_by()`, `mutate()`, and `summarise()` functions. 
Understanding the join functions (especially `inner_join()` and `full_join()` 
will help immensely).

## Piping

The pipe operator `%>%` is a short cut to string any number of operations 
together. So instead of writing a long code worm like
```{r code worm, eval = FALSE}
ungroup(mutate(group_by(select(filter(data, 'Reference' == scenario), region, period, value), region), value = cumsum(value)))
```
that you would have to read from the inside out, you can write each function 
call on a separate line, greatly improving readability as data "flows" from top 
to bottom, each function being applied in succession.
```{r readable code, eval = FALSE}
data %>% 
    filter('Reference' == scenario) %>% 
    select(region, period, value) %>% 
    group_by(region) %>% 
    mutate(value = cumsum(value)) %>% 
    ungroup()
```

## Filtering
`filter()` is a more useful and versatile variant of the data frame extraction 
operator `[`. Instead of
```{r extraction, eval = FALSE}
df[df$column.A == value & df$column.B != 'string',]
```
use
```{r filter, eval = FALSE}
df %>% 
    filter(column.A == value,
           column.B == 'string')
```
again, improving readability and decreasing proneness to error. Predicates (the
test) are combined via logical AND, you can use as many of them as you want, 
and there are lots of helper functions available in the `dplyr` package.

## Selecting
`select()` selects columns from a data frame. So if you only need column A and 
B, do
```{r select A and B, eval = FALSE}
df %>% 
    select(A, B)
```
If you need all but C, do
```{r select all but C, eval = FALSE}
df %>% 
    select(-C)
```
You can rename columns on the way
```{r select and rename, eval = FALSE}
df %>% 
    select(A, B = not.C)
```

## Grouping
`group_by()` subdivides a data frame into groups, such that so-called _window_ 
functions operate only on the items within a group. If you, for example, want 
to calculate the cumulated energy consumption over time for different 
scenarios, you need to group by scenario first.
```{r grouping, eval = FALSE}
df %>% 
    group_by(scenario) %>% 
    mutate(value = cumsum(value)) %>% 
    ungroup()
```

## Mutating
`mutate()` is the low-level function for modifying existing or creating new 
columns. Its output has the same number of rows as its input.
```{r mutating, eval = FALSE}
df %>% 
    group_by(scenario) %>% 
    mutate(value = cumsum(value)) %>% 
    ungroup()
```

## Summarising
`summarise()` is the low-level function for aggregating variables. Its output 
has _one_ row for every group. All ungrouped columns are dropped.
```{r summarising, eval = FALSE}
df %>% 
    group_by(scenario) %>% 
    summarise(value = mean(value)) %>% 
    ungroup()
```


# Load Data
`read.quitte()` reads `.mif`-files and converts them into five-dimensional long 
format data frames with columns `model`, `scenario`, `region`, 
`variable`/`unit`, `period`, and `value`.

```{r load data}
base.path <- 'some/directory/'

data.files <- c(
    'REMIND_generic_r7552c_1p5C_Def-rem-5.mif',
    'REMIND_generic_r7552c_1p5C_UBA_Sust-rem-5.mif',
    'REMIND_generic_r7552c_2C_Def-rem-5.mif',
    'REMIND_generic_r7552c_2C_UBA_Sustlife-rem-5.mif',
    'REMIND_generic_r7552c_REF_Def05-rem-5.mif',
    # 'REMIND_generic_r7552c_REF_Def05-rem-5_tab.mif',
    NULL)
```

> **Tip!** Keep your data input formatted in a way that is easy to read and to 
> modify. Separate the directory from the path if there are many files in the 
> same directory. Having each path (or other item) on a separate line (and 
> ending with `NULL`) makes it easy to comment out individual files during 
> development or debugging.

```{r load data fake, eval = FALSE}
data <- read.quitte(paste0(base.path, data.files))
```

This is what you will do normally. Since the example data is included with the 
`quitte` package, we load the data differently here, which you can ignore.

```{r load data for realz}
(data <- read.quitte(system.file('extdata', data.files, package = 'quitte')))
```

The same data is also available as a data object in the `quitte` package in case 
you just want to try something out or give an example for something.

```{r quitte example data}
quitte_example_data
```


# Trim Data
If you know which data you will need you can trim down the data frame early, 
reducing memory needs and increasing processing speed.
```{r trim data}
scenarios <- inline.data.frame(
    'scenario;                       scen.name',
    'r7552c_REF_Def05-rem-5;         Reference',
    # 'r7552c_1p5C_Def-rem-5;          1.5°C-def',
    'r7552c_2C_Def-rem-5;            2°C',
    'r7552c_1p5C_UBA_Sust-rem-5;     1.5°C-sust',
    # 'r7552c_2C_UBA_Sustlife-rem-5;   2°C-sust',
    NULL)

tmax <- 2100
```

`inline.data.frame()` is a wrapper function for entering tabular data in an 
easy-to-read fashion.

```{r trim data 2}
data <- data %>% 
    filter(scenario %in% scenarios$scenario,
           tmax >= period)
```


# Use Pretty Names
`replace_column()` can be used to replace a column with hard-to-read or ugly 
entries with more pleasant ones. This is quite useful for preparing data for 
plots or tables.

```{r replace scenario names}
(data <- data %>% 
    replace_column(scenarios, scenario, scen.name) %>% 
     order.levels(scenario = scenarios$scen.name))
```

`order.levels()` arranges factor levels (e.g. scenario names) in a specific 
order, thus controlling the order with which items appear in plots.


# Sum Things Up
`sum_total()` sums values up across regions

```{r sum regions}
data %>% 
    filter('Reference' == scenario, 
           'Consumption' == variable, 
           2050 == period, 
           region %in% c('CHN', 'IND', 'JPN', 'OAS')) %>% 
    sum_total(group = region, name = 'all Asia')
```

or across other dimensions

```{r sum variables}
data %>% 
    filter('2°C' == scenario, 
           'RUS' == region,
           2035 == period,
           variable %in% paste0('SE|Heat|', c('Coal', 'Gas'))) %>% 
    sum_total(variable, name = 'SE|Heat|fossil')
```


# Add New Variables
`calc_addVariable()` calculates new variables from existing ones, using generic 
formulas. Variable names, that are not valid R identifiers -- basically 
everything inside a `.mif`-file, needs to be escaped using backticks ("`").

```{r add per-capita FE use}
data %>% 
    filter('EUR' == region,
           2025 == period,
           variable %in% c('PE', 'Population')) %>% 
    calc_addVariable(
        # EJ/yr / million / (3.6e-9 EJ/MWh) * 1e6 1/million = MWh/yr
        '`PE per capita`' = 'PE / Population / 3.6e-3',
        units = 'MWh/year')
```

> **Tip!** Note down all unit conversions you do where you do them, and include 
> the calculation. This will save you hours of debug-time in the future. 
> (Trust me.)

It is also possible to only keep the newly calculated values for further 
manipulation.

```{r add PE carrier intensity of GDP}
data %>% 
    filter('OAS' == region,
           2070 == period,
           'Reference' != scenario) %>% 
    calc_addVariable(
        # EJ/yr / $bn/yr * 1e-15 kJ/EJ * 1e9 $/$bn = kJ/$
        '`PE|Coal|GDP intensity`'       = '`PE|Coal`       / `GDP|PPP` * 1e6',
        '`PE|Oil|GDP intensity`'        = '`PE|Oil`        / `GDP|PPP` * 1e6',
        '`PE|Gas|GDP intensity`'        = '`PE|Gas`        / `GDP|PPP` * 1e6',
        '`PE|Nuclear|GDP intensity`'    = '`PE|Nuclear`    / `GDP|PPP` * 1e6',
        '`PE|Biomass|GDP intensity`'    = '`PE|Biomass`    / `GDP|PPP` * 1e6',
        '`PE|Hydro|GDP intensity`'      = '`PE|Hydro`      / `GDP|PPP` * 1e6',
        '`PE|Geothermal|GDP intensity`' = '`PE|Geothermal` / `GDP|PPP` * 1e6',
        '`PE|Wind|GDP intensity`'       = '`PE|Wind`       / `GDP|PPP` * 1e6',
        '`PE|Solar|GDP intensity`'      = '`PE|Solar`      / `GDP|PPP` * 1e6',
        units = 'kJ/$',
        only.new = TRUE)
```


# Net Present Value of a Time Series

`calcCumulatedDiscount()` calculates the net present value of a time series.

```{r calc NPV}
data %>% 
    filter('Reference' != scenario,
           'World' == region) %>% 
    calcCumulatedDiscount(nameVar = 'Policy Cost|Consumption Loss') %>% 
    select(-model, -region, -unit) %>% 
    spread(scenario, value)
```


# Calculate Sample Quantiles

`calc_quantiles()` makes it easy to produce sample quantiles corresponding 
given probabilities from data frames.

```{r calculate sample quantiles}
(data.plot <- data %>% 
    filter('World' != region,
           2040 == period) %>% 
    calc_addVariable(
        # MtCO2/$bn * 1e9 kg/Mt * 1e-9 $/$bn = kgCO2/$
        '`CO2 intensity`' = '`Emi|CO2` / `GDP|PPP`',
        units = 'kGCO2/US$2005',
        only.new = TRUE) %>% 
    group_by(model, scenario, variable, unit, period) %>% 
    calc_quantiles() %>% 
    spread(quantile, value))
```

This is quite useful for plotting uncertainty ranges.

```{r plot_sample_quantities}
ggplot() +
    geom_col(
        data = data %>% 
            filter('World' == region,
                   2040 == period) %>% 
            calc_addVariable(
                # MtCO2/$bn * 1e9 kg/Mt * 1e-9 $/$bn = kgCO2/$
                '`CO2 intensity`' = '`Emi|CO2` / `GDP|PPP`',
                units = 'kGCO2/US$2005',
                only.new = TRUE),
        mapping = aes(x = scenario, y = value, fill = scenario)) +
    scale_fill_discrete(guide = 'none') +
    geom_boxplot(
        data = data.plot,
        mapping = aes(
            x = scenario, 
            ymin = q0, lower = q25, middle = q50, upper = q75, ymax = q100),
        stat = 'identity',
        width = 0.5)
```

It can also be done for arbitrary quantiles

```{r calculate different sample quantiles}
data %>% 
    filter('World' != region,
           2040 == period) %>% 
    calc_addVariable(
        # MtCO2/$bn * 1e9 kg/Mt * 1e-9 $/$bn = kgCO2/$
        '`CO2 intensity`' = '`Emi|CO2` / `GDP|PPP`',
        units = 'kGCO2/US$2005',
        only.new = TRUE) %>% 
    group_by(model, scenario, variable, unit, period) %>% 
    calc_quantiles(probs = c(low = 1/3, high = 2/3, 'very high' = 4/5)) %>% 
    spread(quantile, value)
```


# Interpolate Missing Periods
`interpolate_missing_periods()` adds missing periods to a data frame and 
interpolates missing values. It uses either _linear_ or _spline_ interpolation 
and can extend missing data before/after the first/last period in the original 
data. 

```{r interpolate missing periods}
data.example <- tribble(
    ~x,   ~value,
    2010,    0.1,
    2015,    0.7,
    2020,    0.9,
    2035,    0.6,
    2040,    0.3)

data.plot <- bind_rows(
    data.example %>% 
        interpolate_missing_periods(
            x = seq(2005, 2045, 1), expand.values = TRUE, 
            method = 'linear') %>% 
        rename(linear = value) %>% 
        gather(interpolation, value, -x),
    
    data.example %>% 
        interpolate_missing_periods(
            x = seq(2005, 2045, 1), expand.values = TRUE, method = 'spline') %>% 
        rename(spline = value) %>% 
        gather(interpolation, value, -x)
)
```

Be careful when using this, as the results may be nonsensical.

```{r plot_interpolations}
ggplot(mapping = aes(x = x, y = value)) +
    geom_line(data = data.plot, mapping = aes(colour = interpolation)) +
    geom_point(data = data.example) +
    coord_cartesian(ylim = c(0, 1))
```

In this case, values below 0 or might be meaningless. So use extension with 
care.


# Plotting with `mip`

Data in `quitte` format is easily passed to the plot functions in the `mip` 
package.

```{r area_plot_with_mip}
data %>% 
    filter('World' == region,
           grepl('^PE\\|', variable)) %>% 
    mip::mipArea(x = .) +
    facet_wrap(~ scenario)
```

```{r bar_plot_with_mip}
data %>% 
    filter(period %in% c(2010, 2030, 2050, 2100),
           grepl('^PE\\|', variable)) %>% 
    mip::mipBarYearData(x = .) +
    facet_wrap(~ region, scales = 'free_y')
```

(Note: You don't have to call the `mip`-functions via the double colon (`::`) 
operator. This is only needed in this vignette, since `quitte` can't import 
`mip`.)


# Plotting 'correct' Bars


If bar plots are used naively (as in: without extra care), the bars for periods 
after 2055 are too narrow and either in the wrong places
```{r bar plot setup}
colours_PE <- c('Coal'       = '#0C0C0C',
                'Oil'        = '#663A00',
                'Gas'        = '#E5E5B2',
                'Nuclear'    = '#FF33FF',
                'Hydro'      = '#191999',
                'Biomass'    = '#005900',
                'Geothermal' = '#E51900',
                'Wind'       = '#337FFF',
                'Solar'      = '#FFCC00')

data_plot <- quitte_example_data %>%
            filter('r7552c_1p5C_Def-rem-5' == scenario,
                   'EUR' == region,
                   grepl('^PE\\|', variable),
                   2100 >= period) %>%
            mutate(variable = sub('^PE\\|', '', variable)) %>%
            order.levels(variable = rev(names(colours_PE)))
```

```{r plot naive bars 2}
ggplot() +
    geom_col(
        data = data_plot,
        mapping = aes(x = factor(period), y = value, fill = variable)) +
    scale_fill_manual(values = colours_PE,
                      name = NULL) +
    labs(x = NULL, y = 'EJ/yr')
```
or spread out with large gaps in between.
```{r plot naive bars 1}
ggplot() +
    geom_col(
        data = data_plot,
        mapping = aes(x = period, y = value, fill = variable)) +
    scale_fill_manual(values = colours_PE,
                      name = NULL) +
    scale_x_continuous(breaks = unique(remind_timesteps$period),
                       name = NULL) +
    labs(y = 'EJ/yr')
```

In both cases, the visual representation of the data is distorted, since the are 
of the bars should correspond to the integral of the variable over time. (3 EJ 
over 10 years should result in a bar with twice the area than 3 EJ over 5 
years.)

To correct this, you can use the function `add_remind_timesteps_columns`, that 
adds two columns to the data frame, `xpos` and `width`, which can be used to 
create a 'correct' bar plot:
```{r correct barplot, warning = FALSE}
ggplot() +
    geom_col(
        data = data_plot %>%
            add_remind_timesteps_columns(gaps = 0.1),
        mapping = aes(x = xpos, width = width, y = value, fill = variable)) +
    scale_fill_manual(values = colours_PE,
                      name = NULL) +
    scale_x_continuous(breaks = unique(remind_timesteps$period),
                       name = NULL) +
    labs(y = 'EJ/yr')
```
